---
title: "Data Cleaning and Transformation"
format: html
editor: visual
---

## Loading Required Libraries

```{r}
# Suppress package messages
library(tidyverse)
```

```{r}
# Load data from a compressed tab-separated file
ncbi_data <- read_tsv("../data/01_ncbi_data.tsv.gz")
```

```{r}
# Extracting patient characteristics
ncbi_characteristics_dirty <- ncbi_data |> 
  as_tibble() |> 
  filter(col_1 == "ID_REF" | 
         col_1 == "!Sample_characteristics_ch1")

# Extracting gene and gene expression data
ncbi_genes_dirty <- ncbi_data |> 
  as_tibble() |> 
  filter(col_1 == "ID_REF" | 
         str_detect(col_1, "ILMN"))

# Preview extracted data (raw and unprocessed)
ncbi_characteristics_dirty
ncbi_genes_dirty
```

# Placeholder for data transformation from long to wide format
# Uncomment and implement pivot_wider if needed
# ncbi_data_wide <- ncbi_characteristics_dirty |> 
#   pivot_wider(names_from = col_1, values_from = everything())

# Renaming columns for clarity
ncbi_characteristics_clean <- ncbi_characteristics_dirty |> 
  rename(Sample_ID = ID_REF)

# Preview cleaned characteristics data
ncbi_characteristics_clean


## Extracting relevant data from NCBI Data

First we extract all data we want to analyze from the data set

```{r}
# Extracting relavant data regarding characteristics of the patients
ncbi_characteristics_dirty <- ncbi_data |> 
  as_tibble() |> 
  filter(col_1 == "ID_REF" | 
        col_1 == "!Sample_characteristics_ch1")

# Extracting relavant data regarding genes and gene expression 
ncbi_genes_dirty <- ncbi_data |> 
  as_tibble() |> 
  filter(col_1 == "ID_REF" |
    str_detect(col_1,"ILMN"))

# View data (dirty) 
ncbi_characteristics_dirty
ncbi_genes_dirty


```

## Data Wrangling - Characteristics

```{r}


# Transforming from long to wide format
#ncbi_data_wide <- 

#ncbi_data_dirty |> 
  #pivot_wider(names_from = col_1, values_from = everything())


# Renaming
ncbi_data_dirty |> 
  rename(Sample_ID = ID_REF)
```

## Data Wrangling - Genes

```{r}

```
