---
title: "Data Cleaning and Transformation"
format: html
editor: visual
---

## Loading Required Libraries

```{r}
# Suppress package messages
library(tidyverse)
```

## Extracting Sample IDs

```{r}
# Inspect the dataset systematically to locate the actual sample IDs
print(head(ncbi_data, 10))  # Inspect the first 10 rows to locate where sample IDs like GSM451162 are stored
sample_id_row <- ncbi_data[which(ncbi_data$col_1 == "ID_REF"), -1]  # Adjust "ID_REF" if needed to match the key identifier
sample_ids <- as.character(sample_id_row)

# Debug: Print extracted sample IDs
print(sample_ids)
```

## Extracting and Filtering Relevant Data

```{r}
# Extract patient characteristics data
ncbi_characteristics_dirty <- ncbi_data %>%
  as_tibble() %>%
  filter(col_1 == "!Sample_characteristics_ch1") %>%
  select(-col_1)  # Drop the first column

# Extract gene expression data
ncbi_genes_dirty <- ncbi_data %>%
  as_tibble() %>%
  filter(str_detect(col_1, "^ILMN"))  # Filter rows with gene IDs (e.g., starting with "ILMN")

# Debug: View the filtered data
print(head(ncbi_characteristics_dirty))
print(head(ncbi_genes_dirty))
```

## Tidying Patient Characteristics Data

```{r}
# Transpose and tidy the characteristics data
tidy_characteristics <- ncbi_characteristics_dirty %>%
  pivot_longer(cols = everything(), names_to = "Sample", values_to = "Characteristic") %>%
  separate(Characteristic, into = c("Variable", "Value"), sep = ": ", extra = "merge") %>%
  pivot_wider(names_from = Variable, values_from = Value) %>%
  mutate(Sample_ID = sample_ids[1:nrow(.)]) %>%  # Align sample IDs
  relocate(Sample_ID)  # Move Sample_ID to the first column

# Drop the "Sample" column and remove rows with "not available"
tidy_characteristics <- tidy_characteristics %>%
  select(-Sample) %>%  # Drop the "Sample" column
  filter(across(everything(), ~ !str_detect(., "not available"), .names = "check"))  # Remove rows with "not available"

# Debug: Verify the reshaped characteristics data
print(head(tidy_characteristics))
```

## Tidying Gene Expression Data

```{r}
# Transpose and tidy the gene expression data
tidy_genes <- ncbi_genes_dirty %>%
  pivot_longer(cols = -col_1, names_to = "Sample", values_to = "Expression") %>%
  pivot_wider(names_from = col_1, values_from = Expression) %>%
  mutate(Sample_ID = sample_ids[1:nrow(.)]) %>%  # Align sample IDs
  relocate(Sample_ID)  # Move Sample_ID to the first column

# Drop the "Sample" column and remove rows with "not available"
tidy_genes <- tidy_genes %>%
  select(-Sample) %>%  # Drop the "Sample" column
  filter(across(everything(), ~ !str_detect(., "not available"), .names = "check"))  # Remove rows with "not available"

# Debug: Verify the reshaped gene data
print(head(tidy_genes))
```

## Cleaning Column Names

```{r}
# Standardize column names for characteristics data
colnames(tidy_characteristics) <- colnames(tidy_characteristics) %>%
  str_replace_all("\\s+", "_") %>%          # Replace spaces with underscores
  str_replace_all("\\.+", "_") %>%          # Replace dots with underscores
  str_replace_all("\\(|\\)", "") %>%        # Remove parentheses
  str_to_lower()                            # Convert to lowercase

# Standardize column names for gene data
colnames(tidy_genes) <- colnames(tidy_genes) %>%
  str_replace_all("\\s+", "_") %>%          # Replace spaces with underscores
  str_replace_all("\\.+", "_") %>%          # Replace dots with underscores
  str_replace_all("\\(|\\)", "") %>%        # Remove parentheses
  str_to_lower()                            # Convert to lowercase

# Debug: Verify cleaned column names
print(colnames(tidy_characteristics))
print(colnames(tidy_genes))
```

## Saving the Cleaned Datasets

```{r}
# Preview the final cleaned datasets
print(tidy_characteristics)
print(tidy_genes)

# Save the cleaned datasets to CSV files
write.csv(tidy_characteristics, "tidy_cleaned_characteristics.csv", row.names = FALSE)
write.csv(tidy_genes, "tidy_cleaned_genes.csv", row.names = FALSE)

```

## Merging Patient Characteristics and Gene Expression Data

```{r}
# Merge the two datasets by `sample_id`
merged_data <- tidy_characteristics %>%
  left_join(tidy_genes, by = "sample_id")  # Perform the merge on `sample_id`

# Debug: Verify the merged data
print(head(merged_data))

# Save the merged dataset to a CSV file
write.csv(merged_data, "merged_cleaned_data.csv", row.names = FALSE)

# Optional: Print final column names to verify structure
#print(colnames(merged_data))

```

# Placeholder for data transformation from long to wide format

# Uncomment and implement pivot_wider if needed

# ncbi_data_wide \<- ncbi_characteristics_dirty \|\>

# pivot_wider(names_from = col_1, values_from = everything())

# Renaming columns for clarity

ncbi_characteristics_clean \<- ncbi_characteristics_dirty \|\> rename(Sample_ID = ID_REF)

# Preview cleaned characteristics data

ncbi_characteristics_clean

## Extracting relevant data from NCBI Data

First we extract all data we want to analyze from the data set

```{r}
# Extracting relavant data regarding characteristics of the patients
ncbi_characteristics_dirty <- ncbi_data |> 
  as_tibble() |> 
  filter(col_1 == "ID_REF" | 
        col_1 == "!Sample_characteristics_ch1")

# Extracting relavant data regarding genes and gene expression 
ncbi_genes_dirty <- ncbi_data |> 
  as_tibble() |> 
  filter(col_1 == "ID_REF" |
    str_detect(col_1,"ILMN"))

# View data (dirty) 
ncbi_characteristics_dirty
ncbi_genes_dirty


```

## Data Wrangling - Characteristics

```{r}


# Transforming from long to wide format
#ncbi_data_wide <- 

#ncbi_data_dirty |> 
  #pivot_wider(names_from = col_1, values_from = everything())


# Renaming
ncbi_data_dirty |> 
  rename(Sample_ID = ID_REF)
```

## Data Wrangling - Genes

```{r}

```
